---
title: "hw03-Jieun-Hwang.Rmd"
author: "JIEUN HWANG"
date: "10/10/2017"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
teams <- read.csv(file="/Users/jieunhwang/stat133/stat133-hws-fall17/hw03/data/nba2017-teams.csv")
```


# Basic Ranking 

## NBA Teams ranked by Total Salary
```{r}
team_salary <- teams %>% select(team,salary) %>% arrange(salary)
team_salary_n <- 1:nrow(teams)

ggplot(team_salary , aes(x=reorder(team,team_salary_n), y=salary)) +
  geom_bar(stat='identity')+
  coord_flip()+
  xlab('Team')+
  ylab('Salary (in millions)')+
  ggtitle("NBA Teams ranked by Total Salary")+
  geom_hline(aes(yintercept=mean(salary)), color = "red")
```

## NBA Teams ranked by Total Points 
```{r}
team_points <- teams %>% select(team,points) %>% arrange(points)
team_points_n <- 1:nrow(teams)

ggplot(team_points , aes(x=reorder(team,team_points_n), y=points)) +
  geom_bar(stat='identity')+
  coord_flip()+
  xlab('Team')+
  ylab('Total Points')+
  ggtitle("NBA Teams ranked by Total Points")+
  geom_hline(aes(yintercept=mean(points)), color = "red")
```

## NBA Teams ranked by Total Efficiency  
```{r}  
team_efficiency <- teams %>% select(team,efficiency) %>% arrange(efficiency)
team_efficiency_n <- 1:nrow(teams)

ggplot(team_efficiency, aes(x=reorder(team,team_efficiency_n), y=efficiency)) +
  geom_bar(stat='identity')+
  coord_flip()+
  xlab('Team')+
  ylab('Total Efficiency')+
  ggtitle("NBA Teams ranked by Total Efficiency")+
  geom_hline(aes(yintercept=mean(efficiency)), color = "red")
```  

# descriptions of the obtained rankings 

1. NBA Teams ranked by Total Salary: max(CLE), min(PHI)
2. NBA Teams ranked by Total Points: max(GSW), min(SAC)
3. NBA Teams ranked by Total Efficiency: max(CLE), min(ORL)



## Principal Components Analysis (PCA)
```{r}
 pca_dat <- data.frame(
          teams$points3,
          teams$points2,
          teams$free_throws,
          teams$off_rebounds,
          teams$def_rebounds,
          teams$assists,
          teams$steals,
          teams$blocks,
          teams$turnovers,
          teams$fouls)
  
 matrix_teams <- as.matrix(pca_dat)
 PCA <- prcomp(matrix_teams, scale. = TRUE)
 
```
 
# Create a data frame with the eigenvalues

```{r}
 eigs <- data.frame(
 eigenvalue = round(PCA$sdev^2,4),
 proportion = round( PCA$sdev^2/ sum(PCA$sdev^2),4),
 cumprop = cumsum(round(PCA$sdev^2/ sum(PCA$sdev^2),4))
 )
 
 eigs
```
 
 
 
# Use the first two PCs to get a scatterplot of the teams

```{r}
# create data frame and find PC1 and PC2

pc_data <- data.frame(PCA$x)
pc_data <- cbind(pc_data,team = teams$team)
pc_data

PC1 = round(pc_data[,1],3)
PC2 = round(pc_data[,2],3)
```

```{r}
# create ggplot

ggplot(data = pc_data) + 
geom_text(aes(x=PC1, y = PC2),label = teams$team)+
xlab('PC1')+
ylab('PC2')+
ggtitle("PCA plot (PC1 and PC2)")

```


# Interpret PC

```{r}
cor(PC1,PC2)
cor(PC1,pc_data$PC3)
cor(pc_data$PC4,pc_data$PC5)
PCA$rotation

```

# Index based on PC1

```{r}
# create data frame 

pc1 = PCA$x[,1]
transf_score <- 100 * (pc1 - min(pc1))/(max(pc1)-min(pc1))

pc1_data <- data.frame(team = teams$team,pc1,transf_score)
pc1_data <- pc1_data %>% arrange(transf_score)
```

```{r}
# create bar chart
ggplot(data = pc1_data, aes(x = reorder(team, transf_score), y = transf_score)) +
coord_flip() +
geom_bar(stat='identity')+
xlab('Team')+
ylab('First PC (scaled from 0 to 100)')+
ggtitle("NBA Teams ranked by scaled PC1")
```
  
# A brief description of the PC1 index to rank the teams

* max(GSW), min(SAC)

# Comments and Reflections
* I spent a lot of time understanding PCA which I've never heard before. But It was interesting by using ggplot and   matrix. Using R scripts and working project was my first experience with coding. Personally, PCA and               ggplot(reverse function) were time consuming parts for me. 
 