---
title: "hw01"
author: "JIEUN HWANG"
date: "9/23/2017"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# data

```{r}
# load data file
load("data/nba2017-salary-points.RData")

# list the available objects
ls()
```

# 1

```{r}
salary_m <-round(salary/1000000, digits=2)
salary_m

experience = replace(experience, experience=="R", 0)
experience
experience <- as.integer(experience) 
class(experience)

```


```{r}
position <- as.factor(position)
class(position)

levels(position) <- c("C"="Center","PF"="power_fwd","PG"="point_guard","SF"="small_fwd","SG"="Shoot_guard")
table(position)
```

# 2

```{r}
plot(points,salary_m,col= "orange", main ="Scatterplot of Points and Salary",xlab="Points", ylab="Salary (in millions)")
```

# 3

```{r}
num_p <- length(points)
mean_of_points <- 1/num_p*sum(points)
mean_of_points

num_s <- length(salary_m)
mean_of_salary_m <- 1/num_s*sum(salary_m)
mean_of_salary_m

var_of_points <- 1/(num_p-1)*sum((points-mean_of_points)^2)
var_of_points

var_of_salary_m <- 1/(num_s-1)*sum((salary_m-mean_of_salary_m)^2)
var_of_salary_m

sd_points <- sqrt(var_of_points)
sd_points

sd_salary_m <- sqrt(var_of_salary_m)
sd_salary_m

cov_p_s <- 1/(num_p-1)*sum((points-mean_of_points)*(salary_m-mean_of_salary_m))
cov_p_s

cor_p_s <- cov_p_s/(sd_points*sd_salary_m)
cor_p_s
```

# 4

```{r}
#slope term
slope <- (cor_p_s)*(sd_salary_m/sd_points)
slope

#intercept term
intercept <- mean_of_salary_m - (slope * mean_of_points)
intercept

# y- hat (the vector of predicted values)
y_hat <- intercept + (slope*points)
y_hat

summary(y_hat)

# inline code for regression equation 
slope1 <- cor(points,salary_m)*sd(salary_m)/sd(points)
intercept0 <- mean(salary_m)-(slope1*mean(points))
regr_equation <- intercept0 + (slope1*points)
regr_equation
summary(regr_equation)
```

# how do you interpret the slope coeffcient?
# it represents the rate of change of salary_m (Y) as a function of changes in points(X).

# how you interpret the intercept term?
# The intercept is the expected mean value of salary_m (Y) when all points(X)=0.

```{r}
# what is the predicted salary for a player that score?
predict_0s <- intercept0 + (slope1*0)
predict_0s

predict_100s <- intercept0 +(slope1*100)
predict_100s

predict_500s <- intercept0 +(slope1*500)
predict_500s

predict_1000s <- intercept0 +(slope1*1000)
predict_1000s

predict_2000s <- intercept0 +(slope1*2000)
predict_2000s

```

# 5

```{r}
plot(points, salary_m, col ="orange", main = "Regression and Lowess lines",xlab="Points", ylab="Salary (in millions)")
abline(a = intercept0, b =slope1,col="blue")
lines(lowess(points,salary_m),col="red")
text(2500,30,"lowess", col ="red")
text(2400,19,"regression", col ="blue")
```

# 6

```{r}
R <- salary_m - y_hat
summary(R)

RSS <- sum(R^2)
RSS

TSS <- sum((salary_m - mean(salary_m)^2))
TSS

R2coefficient <- 1-(RSS/TSS)
R2coefficient
```

#7

```{r}
plot(experience, salary_m,xlab="Points", ylab="Salary (in millions)",main ="Scatterplot with lowess smooth",col ="pink")
lines(lowess(experience,salary_m),col="blue")
require(scatterplot3d)
scatterplot3d(points,experience,salary_m,main="3D Scatterplot")

boxplot(salary_m ~ position, xlab="Position")

```

* scatterplot with lowess smooth : result is bell shape. It seems there is no strong relation between experience and salary.
* 3D scatterplot : more points, more salary but not really strong connection with experience 
* boxplot : not really strong relation between position and salary.
* does experience seem to be related with salary? : Not really, because after 5 years experience, the line of lowess is decreasing.
* does position seem to be related with salary? : Not really, because the median is similar for each position 


#8

* Honestly, everything was difficult for me. I finished homework alone and it tooks more than 6 hours. I really needed a help escpecially boxplot problem. This part made me feel frustrated. I realized I made a mistake of #1. That's why it didn't work when I made boxplot. I should work with the group next time. It is a challenge!


